---
title: "SYNFloodPDA"
authors: "Taulant, Amir, Jatin, Carter"
date: "2025-03-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import Libraries
```{r}
library(dplyr)
library(ggplot2)
```

# Import raw dataset

# Import Balanced dataset
```{r}
# Import syn dataset 
syn_dt <- read.csv("Balanced-SYN-V2.csv")

# Explor the dataset
str(syn_dt)

```

## Data Preprocessing

```{r}
# Select relevant variables with dyplr library

syn_selected_dt <- syn_dt %>% select('SYN.Flag.Count',
                                     'Total.Fwd.Packets',
                                     'Total.Backward.Packets',
                                     'Flow.Duration',
                                     'Flow.Packets.s',
                                     'Flow.Bytes.s',
                                     'Fwd.Packet.Length.Mean',
                                     'Bwd.Packet.Length.Mean',
                                     'Bwd.IAT.Mean',
                                     'ACK.Flag.Count',
                                     'Active.Mean',
                                     'Label',
                                     'Inbound')
```

```{r}
# Use summary function to explore selected dataset
summary(syn_selected_dt)
```


## Data Cleaning

```{r}
# Removing NA values
syn_selected_dt <- na.omit(syn_selected_dt)

```

```{r}
# Removing infinitty values
syn_selected_dt[syn_selected_dt == "Inf"] <- NA
syn_selected_dt <- na.omit(syn_selected_dt)
```

```{r}
syn_selected_dt$Label <- as.factor(syn_selected_dt$Label)
```

```{r}
str(syn_selected_dt)
```
```{r}
str(syn_selected_dt$Flow.Duration)
```


## EDA

```{r}
# Create a table with flag count 
flag_count_table <- table(syn_selected_dt$SYN.Flag.Count)
flag_count_table

```

```{r}
# Create a table w;ith Total BWD Packets
Total_Backward_Packets_table <- table(syn_selected_dt$Total.Backward.Packets)
Total_Backward_Packets_table
boxplot(log(Total_Backward_Packets_table))

```


```{r}
# Create a table with Total FWD Packet
Total_Fwd_Packet_table <- table(syn_selected_dt$Total.Fwd.Packets)
Total_Fwd_Packet_table
boxplot(log(Total_Fwd_Packet_table))
```

```{r}
# Create a table with Flow Duration
#Flow_Duration_table <- table(syn_selected_dt$Flow.Duration)
#Flow_Duration_table

hist(syn_selected_dt$Flow.Duration)
boxplot(log(syn_selected_dt$Flow.Duration))
```

```{r}
# Create a table with Flow labels
Labels_table <- table(syn_selected_dt$Label)
Labels_table
boxplot(log(Labels_table))
```

```{r}
boxplot(
  log10(syn_selected_dt$SYN.Flag.Count + 1),
  log10(syn_selected_dt$Total.Fwd.Packets + 1),
  log(syn_selected_dt$Total.Backward.Packets + 1),
  log(syn_selected_dt$Flow.Duration + 1),
  log2(syn_selected_dt$Flow.Packets.s + 1),
  log2(syn_selected_dt$Flow.Bytes.s + 1),
  log(syn_selected_dt$Fwd.Packet.Length.Mean + 1),
  log(syn_selected_dt$Bwd.Packet.Length.Mean + 1),
  log(syn_selected_dt$Bwd.IAT.Mean + 1),
  log(syn_selected_dt$ACK.Flag.Count + 1),
  names = c(
    "SYN.Flag.Count", 
    "Total.Fwd.Packets", 
    "Total.Backward.Packets",
    "Flow.Duration",
    "Flow.Packets.s",
    "Flow.Bytes.s",
    "Fwd.Packet.Length.Mean",
    "Bwd.Packet.Length.Mean",
    "Bwd.IAT.Mean",
    "ACK.Flag.Count"
  ),
  main = "Combined Boxplots",
  las  = 2 
)

```
```{r}
t.test(SYN.Flag.Count ~ Label, data = syn_selected_dt)

```
```{r}
t.test(Total.Fwd.Packets ~ Label, data = syn_selected_dt)

```
```{r}
t.test(Total.Backward.Packets ~ Label, data = syn_selected_dt)

```
```{r}
t.test(Flow.Duration ~ Label, data = syn_selected_dt)

```

```{r}
t.test(Flow.Packets.s ~ Label, data = syn_selected_dt)

```

```{r}
t.test(Flow.Bytes.s ~ Label, data = syn_selected_dt)

```

```{r}
t.test(Fwd.Packet.Length.Mean ~ Label, data = syn_selected_dt)

```

```{r}
t.test(Bwd.Packet.Length.Mean ~ Label, data = syn_selected_dt)

```

```{r}
t.test(Bwd.IAT.Mean ~ Label, data = syn_selected_dt)

```

```{r}
t.test(ACK.Flag.Count ~ Label, data = syn_selected_dt)

```




### PCA
```{r}
#names(syn_selected_dt)
#str(syn_selected_dt)

# Create the groups
all_dt_ex_lable <- as.data.frame(syn_selected_dt[,-12])

```

```{r}
# Calculate PCA on whole dataset

PCA_results <- prcomp(all_dt_ex_lable , center = T, scale. = T)
```

```{r}
# visualising pca results

### 4.1 calculate the proportion of exaplained variance (PEV) from the std values
PCA_results_var <- PCA_results$sdev^2
PCA_result_PEV <- PCA_results_var / sum(PCA_results_var)
### 4.2 plot the cumulative PEV
opar <- par(no.readonly = TRUE)
plot(
  cumsum(PCA_result_PEV),
  ylim = c(0,1),
  xlab = 'PC',
  ylab = 'cumulative PEV',
  pch = 20,
  col = 'orange'
)
abline(h = 0.8, col = 'red', lty = 'dashed')
par(opar)

### 4.2b barplot of individual PEV (scree plot)
barplot(
  PCA_result_PEV,
  names.arg = paste0("PC", seq_along(PCA_result_PEV)),
  las = 2,
  col = "orange",
  ylab = "Explained Variance",
  xlab = "Principal Components",
  main = "Bar chart of PEV for each PC",
  ylim = c(0, max(PCA_result_PEV) * 1.1)
)

### 4.3 get and inspect the loadings
PCA_result_loadings <- PCA_results$rotation
PCA_result_loadings

### 4.4 generate a biplot for PC1 and PC2
opar <- par(no.readonly = TRUE)
biplot(
  PCA_results,
  scale = 0,
  col = c('grey40','orange')
)
par(opar)


```



so we know its 1 to 6 that we need so we assign only 1 to 6 to the dataframe as a subset of the PCA, then we do Cluster analysis on that. Centres is the amount of clusters, 25 is how many times k means is run. we convert the matrix to a dataframe and then we add the cluster result through "as factor" to the dataframe. 

```{r}
first_six_PC <- PCA_results$x[, 1:6]

kmeans_result <- kmeans(first_six_PC, centers = 4, nstart = 25)
PCA_subset <- as.data.frame(first_six_PC)
PCA_subset$cluster <- as.factor(kmeans_result$cluster)

```

You can only plot 2 PCs and see what clusters they form against eachother to visualise it so here are two examples of that.

```{r}
plot(
  PCA_subset$PC1, PCA_subset$PC2,
  col = PCA_subset$cluster,
  pch = 19,
  xlab = "PC1", ylab = "PC2",
  main = "K-means Clustering on PC1 vs PC2"
)
legend("topright", legend = levels(PCA_subset$cluster), col = 1:3, pch = 19)

```
```{r}
plot(
  PCA_subset$PC3, PCA_subset$PC4,
  col = PCA_subset$cluster,
  pch = 19,
  xlab = "PC3", ylab = "PC4",
  main = "K-means Clustering on PC3 vs PC4"
)
legend("topright", legend = levels(PCA_subset$cluster), col = 1:3, pch = 19)

```



## Save slected dataset

```{r}
# Save the final cleaned dataset
#write.csv(syn_selected_dt , "syn seleceted dt.csv")
```

